Processing CDC hypertension and CDV Mortality disease data from adults in the USA - Processing script

### 2025-05-02 code/eda-code/eda.qmd
Changes for final product

@@

```{r}
#load needed packages. make sure they are installed.
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set saving/loading paths
library(knitr) #for nice tables and data integration
library(corrplot) #for correlation matrix visualization
library(sf) #for geospatial analysis
library(ggpubr) #for advanced visualization tools
library(ggplot2) #for plots
library(gt) #For a tidy table

@@

saveRDS(revised_mortalities_histogram, file = here::here("data","processed-data", "revised_mortalities_histogram.rds"))
revised_mortalities_histogram <- readRDS(here::here("data","processed-data", "revised_mortalities_histogram.rds"))
# Print the mortalities histogram
print(revised_mortalities_histogram)
```

@@

# Optional: Save these for later
saveRDS(age_group_1, here::here("data", "processed-data", "age_group_1.rds"))
saveRDS(age_group_2, here::here("data", "processed-data", "age_group_2.rds"))

@@

# Save new histo variable
saveRDS(hist_age_group_1, here::here("data", "processed-data", "hist_age_group_1.rds"))
#Print histogram
print(hist_age_group_1)

@@

# Save and optionally reload
saveRDS(hist_age_group_2, here::here("data", "processed-data", "hist_age_group_2.rds"))
# Print histogram 2
print(hist_age_group_2)

@@

saveRDS(boxplot_race_sex, file = here::here("data","processed-data", "boxplot_race_sex.rds"))
print(boxplot_race_sex)

boxplot_sex <- ggplot(county_sex, aes(x = sex, y = mortalities)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Mortality Rates by Sex", x = "Sex", y = "Mortality Rate")

saveRDS(boxplot_sex, file = here::here("data","processed-data", "boxplot_sex.rds"))
print(boxplot_sex)

@@

Neatening up the boxplot for race without sex
```{r}
# Relationship between mortality and race
race_plot <- ggplot(county_race, aes(x = race_ethnicity, y = mortalities)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Mortality Rates by Race/Ethnicity", x = "Race/Ethnicity", y = "Mortality Rate") +
  theme(axis.text.x = element_text(size = 10)) +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10))

ggsave(here::here("results", "figures", "mortality_by_race.png"), plot = race_plot)
print(race_plot)

```
This is the same but without sex and the colours for a neater output.

@@

saveRDS(national_age_trend, here::here("data", "processed-data", "national_trend_by_age.rds"))
print(national_age_trend)
```

@@

saveRDS(race_trend_65_plus, here::here("data", "processed-data", "race_trend_65_plus.rds"))
print(race_trend_65_plus)

@@

saveRDS(race_trend_35_64, here::here("data", "processed-data", "race_trend_35_64.rds"))
print(race_trend_35_64)

@@

saveRDS(sex_trend_65_plus, here::here("data", "processed-data", "sex_trend_65_plus.rds"))
print(sex_trend_65_plus)

@@

saveRDS(sex_trend_35_64, here::here("data", "processed-data", "sex_trend_35_64.rds"))
print(sex_trend_35_64)

@@

gtsave(grouped_summary_clean, here::here("results", "tables", "summary_mortality_table.png"))

@@

saveRDS(yearly_means, here::here("data", "processed-data", "yearly_means.rds"))
```

>>>>

This should be the second file to run when reproducing the full analysis for the project.

@@

```{r}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set saving/loading paths
library(knitr) #for nice tables and data integration
library(corrplot) #for correlation matrix visualization
library(sf) #for geospatial analysis
library(ggpubr) #for advanced visualization tools
library(ggplot2) #for plots
library(gt) #For a tidy table
library(stringr)

@@

invisible(saveRDS(revised_mortalities_histogram, file = here::here("results","output", "eda", "revised_mortalities_histogram.rds")))
revised_mortalities_histogram <- readRDS(here::here("results","output", "eda", "revised_mortalities_histogram.rds"))
# Print the mortalities histogram
print(revised_mortalities_histogram)
```

@@

# Save these for later
saveRDS(age_group_1, here::here("data", "processed-data", "age_group_1.rds"))
saveRDS(age_group_2, here::here("data", "processed-data", "age_group_2.rds"))

@@

# Save new histo variable
invisible(saveRDS(hist_age_group_1, here::here("results", "output", "eda", "hist_age_group_1.rds")))
#Print histogram
print(hist_age_group_1)

@@

# Save and optionally reload
invisible(saveRDS(hist_age_group_2, here::here("results", "output", "eda", "hist_age_group_2.rds")))
# Print histogram 2
print(hist_age_group_2)

@@

invisible(saveRDS(boxplot_race_sex, file = here::here("results", "output", "eda", "boxplot_race_sex.rds")))
print(boxplot_race_sex)

Summary Statistics of mean and median:

```{r}
# Summarize mean and median
mortality_summary <- county_race %>%
  group_by(race_ethnicity) %>%
  summarise(
    Mean = round(mean(mortalities, na.rm = TRUE), 1),
    Median = round(median(mortalities, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  rename(`Race/Ethnicity` = race_ethnicity)

# Create gt table
mortality_summary_gt <- mortality_summary %>%
  gt() %>%tab_header(
    title = "County-Level Mortality by Race/Ethnicity (2000–2019)",
    subtitle = "Mean and median mortality rates per 100,000 population") %>%
  cols_label(Mean = html("Mean Mortality<br>(per 100,000)"),
    Median = html("Median Mortality<br>(per 100,000)")) %>%
  cols_align(align = "left", columns = `Race/Ethnicity`) %>%
  cols_align(align = "right", columns = c(Mean, Median)) %>%
  fmt_number(columns = c(Mean, Median), decimals = 1) %>%
  opt_table_font(font = google_font("Roboto")) %>%
  tab_options(table.width = pct(100), column_labels.font.weight = "bold",
    data_row.padding = px(3),heading.align = "center")

# Save table
gtsave(
  mortality_summary_gt,
  filename = here::here("results", "tables", "mortality_summary_by_race.png"))

mortality_summary_gt
```

Boxpplot by sex:
```{r}
boxplot_sex <- ggplot(county_sex, aes(x = sex, y = mortalities)) +
  geom_boxplot(fill = "#4C78A8", alpha = 0.7, outlier.size = 0.7, outlier.alpha = 0.5) +
  labs(
    title = "County-Level Mortality Rates by Sex",
    x = "Sex",
    y = "Mortality per 100,000"
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    axis.text.x = element_text(size = 10, hjust = 0.5))

boxplot_sex

# Save file as png
ggsave(here::here("results", "figures", "mortality_by_sex.png"), plot = boxplot_sex,
  width = 8, height = 6, dpi = 300)

@@

Neatening up the boxplot for race without sex

```{r}
# Plot: Mortality by race (boxplot)
mortality_by_race_plot <- ggplot(county_race, aes(x = race_ethnicity, y = mortalities)) +
  geom_boxplot(fill = "#4C78A8", alpha = 0.7, outlier.size = 0.7, outlier.alpha = 0.5) +
  stat_summary(fun = "mean", geom = "point", shape = 20, size = 2.8, color = "red") +
  labs(
    title = "County-Level Mortality Rates by Race/Ethnicity",
    x = "Race/Ethnicity",
    y = "Mortality per 100,000"
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),     
    axis.title.y = element_text(size = 14),     
    axis.text.x = element_text(size = 12, hjust = 0.5),
    axis.text.y  = element_text(size = 12),)

mortality_by_race_plot

# Save file as png
ggsave(here::here("results", "figures", "mortality_by_race.png"),
  plot = mortality_by_race_plot, width = 8, height = 6, dpi = 300)

This is the same but without sex and the colours for a neater output and inclusion of a dot to represent the mean values.

Boxplot of county-level mortality rates by race/ethnicity in the U.S. (2000–2019).
Red dots represent mean mortality. Although White adults had the highest median, Black adults experienced the highest mean mortality, reflecting greater spread and right-skew in their distribution.

@@

invisible(saveRDS(national_age_trend, here::here("results", "output", "eda", "national_trend_by_age.rds")))
print(national_age_trend)
```

@@

invisible(saveRDS(race_trend_65_plus, here::here("results", "output", "eda", "race_trend_65_plus.rds")))
print(race_trend_65_plus)

@@

invisible(saveRDS(race_trend_35_64, here::here("results", "output", "eda", "race_trend_35_64.rds")))
print(race_trend_35_64)

@@

invisible(saveRDS(sex_trend_65_plus, here::here("results", "output", "eda", "sex_trend_65_plus.rds")))
print(sex_trend_65_plus)

@@

invisible(saveRDS(sex_trend_35_64, here::here("results", "output", "eda", "sex_trend_35_64.rds")))
print(sex_trend_35_64)

@@

invisible(gtsave(grouped_summary_clean, here::here("results", "tables", "summary_mortality_table.png")))

@@

invisible(saveRDS(yearly_means, here::here("results", "output", "eda", "yearly_means.rds")))
```

### 2025-05-02 code/processing-code/processingfile-v1.qmd
Changes for final product

@@

```{r}
skim_summary_hypertens_filtered<-skimr::skim(hypertens_adult_filtered)|> as.data.frame()
kable(skim_summary_hypertens_filtered)
saveRDS(skim_summary_hypertens_filtered, file = here::here("data","processed-data", "skim_summary_hypertens_filtered.rds"))
print(skim_summary_hypertens_filtered)

@@

```{r histogram mortalities}
#Generate a histogram of the mortality rates
mortalities_histogram <- hist(hypertens_adult_filtered$mortalities)
saveRDS(mortalities_histogram, file = here::here("data","processed-data", "mortalities_histogram.rds"))
mortalities_histogram <- readRDS(here::here("data","processed-data", "mortalities_histogram.rds"))
# Print the mortalities histogram
print(mortalities_histogram)
```

>>>>

This should be the first file to run when reproducing the full analysis.

@@

```{r}
skim_summary_hypertens_filtered<-skimr::skim(hypertens_adult_filtered)|> as.data.frame()
kable(skim_summary_hypertens_filtered)
invisible(saveRDS(skim_summary_hypertens_filtered, file = here::here("results", "output", "processing", "skim_summary_hypertens_filtered.rds")))
print(skim_summary_hypertens_filtered)

@@

```{r histogram mortalities}
#Generate a histogram of the mortality rates
mortalities_histogram <- hist(hypertens_adult_filtered$mortalities)
invisible(saveRDS(mortalities_histogram, file = here::here("results","output", "processing", "mortalities_histogram.rds")))
mortalities_histogram <- readRDS(here::here("results", "output","processing", "mortalities_histogram.rds"))
# Print the mortalities histogram
print(mortalities_histogram)
```
